---
title: "analysis"
author: "joshjiang"
date: "8/6/2019"
output: html_document
---

```{r setup,include=FALSE}
library(ggplot2)
library(tidyverse)
library(gridExtra)
```

Problem:

The beta prime vs secondary plots are not in the same scale.

e.g.

```{r, echo=FALSE,warning=FALSE,message=FALSE}

dapg_file_pattern <- "/Users/joshjiang/hakylab/gtex-gwas-analysis/data/dapg_selected_variants/expression/gwas_and_eqtl/DAPG_with_mashr__{tissue}.rds"
phenotype_lst <- readLines("/Users/joshjiang/hakylab/gtex-gwas-analysis/data/phenotype_list.txt")
metadata <- read_tsv("/Users/joshjiang/hakylab/gtex-gwas-analysis/data/gwas_metadata.txt") %>% rename(phenotype=Tag)
tissue <- "Whole_Blood"

phenotype <- "Astle_et_al_2016_Eosinophil_counts"
df <- readRDS(glue::glue(dapg_file_pattern))[[phenotype]]
short_phenotype <- as.character(metadata[metadata$phenotype==phenotype, "new_Phenotype"])
df_2_eqtl <- df %>%
  group_by(tissue, phenotype, gene_id) %>%
  filter(n() == 2) %>%
  mutate(rank = order(order(abs(eqtl_effect_size), decreasing=TRUE))) %>%
  mutate(rank = paste0("rank", rank)) %>%
  mutate(beta_gene=gwas_effect_size/eqtl_effect_size) %>%
  select(tissue, phenotype, gene_id, rank, beta_gene) %>%
  spread(key=rank, value=beta_gene)
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank1 > quantile(df_2_eqtl$rank1, 0.05, na.rm = TRUE) & df_2_eqtl$rank1 < quantile(df_2_eqtl$rank1, 0.95, na.rm = TRUE),]
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank2 > quantile(df_2_eqtl$rank2, 0.05, na.rm = TRUE) & df_2_eqtl$rank2 < quantile(df_2_eqtl$rank2, 0.95, na.rm = TRUE),]

plot_title <- "{gsub(pattern='_', replacement=' ', x=tissue)} - {gsub(pattern='_', replacement=' ', x=short_phenotype)}\nPrimary vs. secondary eQTL" %>% glue::glue()

pp1 <- ggplot(df_2_eqtl, aes(rank1, rank2))
pp1 <- pp1 + geom_point(alpha=.2, size=4)
pp1 <- pp1 + theme_bw(base_size=20)
pp1 <- pp1 + geom_abline(slope=1,intercept=0)
pp1 <- pp1 + stat_density2d(color = "black", contour = TRUE)
pp1 <- pp1 + ggtitle(plot_title)
pp1 <- pp1 + xlab(expression(beta[prim])) + ylab(expression(beta[sec]))
pp1 <- pp1 + xlim(c(-0.06,0.06)) + ylim(c(-0.06,0.06))

phenotype <- "UKB_20002_1094_self_reported_deep_venous_thrombosis_dvt"
df <- readRDS(glue::glue(dapg_file_pattern))[[phenotype]]
short_phenotype <- as.character(metadata[metadata$phenotype==phenotype, "new_Phenotype"])
df_2_eqtl <- df %>%
  group_by(tissue, phenotype, gene_id) %>%
  filter(n() == 2) %>%
  mutate(rank = order(order(abs(eqtl_effect_size), decreasing=TRUE))) %>%
  mutate(rank = paste0("rank", rank)) %>%
  mutate(beta_gene=gwas_effect_size/eqtl_effect_size) %>%
  select(tissue, phenotype, gene_id, rank, beta_gene) %>%
  spread(key=rank, value=beta_gene)
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank1 > quantile(df_2_eqtl$rank1, 0.05, na.rm = TRUE) & df_2_eqtl$rank1 < quantile(df_2_eqtl$rank1, 0.95, na.rm = TRUE),]
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank2 > quantile(df_2_eqtl$rank2, 0.05, na.rm = TRUE) & df_2_eqtl$rank2 < quantile(df_2_eqtl$rank2, 0.95, na.rm = TRUE),]

plot_title <- "{gsub(pattern='_', replacement=' ', x=tissue)} - {gsub(pattern='_', replacement=' ', x=short_phenotype)}\nPrimary vs. secondary eQTL" %>% glue::glue()

pp2 <- ggplot(df_2_eqtl, aes(rank1, rank2))
pp2 <- pp2 + geom_point(alpha=.2, size=4)
pp2 <- pp2 + theme_bw(base_size=20)
pp2 <- pp2 + geom_abline(slope=1,intercept=0)
pp2 <- pp2 + stat_density2d(color = "black", contour = TRUE)
pp2 <- pp2 + ggtitle(plot_title)
pp2 <- pp2 + xlab(expression(beta[prim])) + ylab(expression(beta[sec]))
pp2 <- pp2 + xlim(c(-0.06,0.06)) + ylim(c(-0.06,0.06))

grid.arrange(pp1,pp2,nrow=1)
```

This can be fixed by normalizing the effect sizes (dividing the gwas and eqtl effect sizes by their respective standard deviations) and readjusting the limits:

```{r, echo=FALSE,warning=FALSE,message=FALSE}

phenotype <- "Astle_et_al_2016_Eosinophil_counts"
df <- readRDS(glue::glue(dapg_file_pattern))[[phenotype]]
df['eqtl_effect_size'] <- df['eqtl_effect_size']/sd(df$eqtl_effect_size, na.rm=TRUE)
df['gwas_effect_size'] <- df['gwas_effect_size']/sd(df$gwas_effect_size, na.rm=TRUE)
short_phenotype <- as.character(metadata[metadata$phenotype==phenotype, "new_Phenotype"])
df_2_eqtl <- df %>%
  group_by(tissue, phenotype, gene_id) %>%
  filter(n() == 2) %>%
  mutate(rank = order(order(abs(eqtl_effect_size), decreasing=TRUE))) %>%
  mutate(rank = paste0("rank", rank)) %>%
  mutate(beta_gene=gwas_effect_size/eqtl_effect_size) %>%
  select(tissue, phenotype, gene_id, rank, beta_gene) %>%
  spread(key=rank, value=beta_gene)
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank1 > quantile(df_2_eqtl$rank1, 0.05, na.rm = TRUE) & df_2_eqtl$rank1 < quantile(df_2_eqtl$rank1, 0.95, na.rm = TRUE),]
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank2 > quantile(df_2_eqtl$rank2, 0.05, na.rm = TRUE) & df_2_eqtl$rank2 < quantile(df_2_eqtl$rank2, 0.95, na.rm = TRUE),]

plot_title <- "{gsub(pattern='_', replacement=' ', x=tissue)} - {gsub(pattern='_', replacement=' ', x=short_phenotype)}\nPrimary vs. secondary eQTL" %>% glue::glue()

pp1 <- ggplot(df_2_eqtl, aes(rank1, rank2))
pp1 <- pp1 + geom_point(alpha=.2, size=4)
pp1 <- pp1 + theme_bw(base_size=20)
pp1 <- pp1 + geom_abline(slope=1,intercept=0)
pp1 <- pp1 + stat_density2d(color = "black", contour = TRUE)
pp1 <- pp1 + ggtitle(plot_title)
pp1 <- pp1 + xlab(expression(beta[prim])) + ylab(expression(beta[sec]))
pp1 <- pp1 + xlim(c(-2,2)) + ylim(c(-2,2))

phenotype <- "UKB_20002_1094_self_reported_deep_venous_thrombosis_dvt"
df <- readRDS(glue::glue(dapg_file_pattern))[[phenotype]]
df['eqtl_effect_size'] <- df['eqtl_effect_size']/sd(df$eqtl_effect_size, na.rm=TRUE)
df['gwas_effect_size'] <- df['gwas_effect_size']/sd(df$gwas_effect_size, na.rm=TRUE)
short_phenotype <- as.character(metadata[metadata$phenotype==phenotype, "new_Phenotype"])
df_2_eqtl <- df %>%
  group_by(tissue, phenotype, gene_id) %>%
  filter(n() == 2) %>%
  mutate(rank = order(order(abs(eqtl_effect_size), decreasing=TRUE))) %>%
  mutate(rank = paste0("rank", rank)) %>%
  mutate(beta_gene=gwas_effect_size/eqtl_effect_size) %>%
  select(tissue, phenotype, gene_id, rank, beta_gene) %>%
  spread(key=rank, value=beta_gene)
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank1 > quantile(df_2_eqtl$rank1, 0.05, na.rm = TRUE) & df_2_eqtl$rank1 < quantile(df_2_eqtl$rank1, 0.95, na.rm = TRUE),]
df_2_eqtl <- df_2_eqtl[df_2_eqtl$rank2 > quantile(df_2_eqtl$rank2, 0.05, na.rm = TRUE) & df_2_eqtl$rank2 < quantile(df_2_eqtl$rank2, 0.95, na.rm = TRUE),]

plot_title <- "{gsub(pattern='_', replacement=' ', x=tissue)} - {gsub(pattern='_', replacement=' ', x=short_phenotype)}\nPrimary vs. secondary eQTL" %>% glue::glue()

pp2 <- ggplot(df_2_eqtl, aes(rank1, rank2))
pp2 <- pp2 + geom_point(alpha=.2, size=4)
pp2 <- pp2 + theme_bw(base_size=20)
pp2 <- pp2 + geom_abline(slope=1,intercept=0)
pp2 <- pp2 + stat_density2d(color = "black", contour = TRUE)
pp2 <- pp2 + ggtitle(plot_title)
pp2 <- pp2 + xlab(expression(beta[prim])) + ylab(expression(beta[sec]))
pp2 <- pp2 + xlim(c(-2,2)) + ylim(c(-2,2))

grid.arrange(pp1,pp2,nrow=1)
```

An interesting observation is that the variance of $\beta_{sec}$ is higher than that of $\beta_{prim}$. This is obvious if you zoom out of both of the above graphs.

```{r, echo=FALSE,warning=FALSE,message=FALSE}

pp1 <- pp1 + xlim(c(-4,4)) + ylim(c(-4,4))

pp2 <- pp2 + xlim(c(-4,4)) + ylim(c(-4,4))

grid.arrange(pp1,pp2,nrow=1)
```

